\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[12pt,draft]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
    \setmainfont[]{DejaVu Serif}
    \setmonofont[Mapping=tex-ansi]{DejaVu Sans Mono}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{pmboxdraw}
\usepackage{authblk}
% \usepackage{amsfonts}
% \usepackage{mathpazo}
% \usepackage[no-math]{fontspec}
% \usepackage{fontenc}

% \setromanfont{DejaVu Serif}
% \setmathfont(Pica){DejaVu Math TeX Gyre}
% \setmathrm{DejaVu Serif}
% \renewcommand{\rmdefault}{DejaVu Serif}
% \usepackage{color}   %May be necessary if you want to color links
% \usepackage{hyperref}
% \hypersetup{
%     colorlinks=true, %set true if you want colored links
%     linktoc=all,     %set to all if you want both sections and subsections linked
%     linkcolor=blue,  %choose some color if you want links to stand out
% }

% \setcounter{secnumdepth}{3}

\renewcommand{\Affilfont}{\raggedright\footnotesize}

% \definecolor{CodeColor}{rgb}{0.858, 0.188, 0.478}
\definecolor{CodeColor}{rgb}{.6, .4, .1}

\let\oldtexttt\texttt
\renewcommand{\texttt}[1]{\textcolor{CodeColor}{\oldtexttt{#1}}}

\let\oldverbatim\verbatim
\renewenvironment{verbatim}
    {\color{CodeColor}\begin{oldverbatim}}
    {\end{oldverbatim}}

% \newcommand{\oldtexttt}{\@texttt}
% \renewcommand{\code}{\textcolor{red}}

% \makeatletter  % Magic for using @ in command names
% \renewcommand{\maketitle}{
%     \bgroup
%      \@title
%     \begin{flushleft}
%         \@author
%     \end{flushleft}
%     % \@date
%     \egroup
% }
% \makeatother

\date{}

\begin{document}

\title{Supplementary information for \\ ``OverProt: secondary structure consensus for protein families''}
% \title{OverProt -- Description of methods}

% $if(title)$ $endif$

\author[1,2]{Adam Midlik}
\author[1,2,3]{Ivana Hutařová Vařeková}
\author[1,2]{Jan Hutař}
\author[1,2]{Aliaksei Chareshneu}
\author[4,*]{Karel Berka}
\author[1,2,*]{Radka Svobodová}

\affil[1]{CEITEC -- Central European Institute of Technology, Masaryk University, Kamenice 5, 625 00 Brno, Czech Republic}
\affil[2]{National Centre for Biomolecular Research, Faculty of Science, Masaryk University, Kamenice 5, 625 00 Brno, Czech Republic}
\affil[3]{Faculty of Informatics, Masaryk University, Botanická 68a, 602 00 Brno, Czech Republic}
\affil[4]{Department of Physical Chemistry, Regional Centre of Advanced Technologies and Materials, Faculty of Science, Palacký University, 17. listopadu 1192/12, 771 46 Olomouc, Czech Republic}
\affil[*]{To whom correspondence should be addressed.}
% \date{\today}


\maketitle

% \tableofcontents

\renewcommand*\contentsname{Table of contents}
{
\setcounter{tocdepth}{3}
\tableofcontents
}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

OverProt is a tool for construction and visualization of the secondary
structure consensus for protein families. The consensus produced by
OverProt can be used as a template for annotation of secondary structure
elements by SecStrAnnotator.

OverProt consists of three main parts: the main algorithm
\textbf{OverProt Core} constructs the secondary structure consensus,
\textbf{OverProt Viewer} visualizes the consensus, and \textbf{OverProt
Server} presents the results on the web and allows user-defined
computations.

\hypertarget{terminology}{%
\section{Terminology}\label{terminology}}

\begin{itemize}
\item
  \textbf{Protein structure} -- a set of atoms with assigned 3D
  coordinates. A structure consists of one ore more \textbf{chains}. A
  chain is a sequence of \textbf{residues}, which consist of the
  individual \textbf{atoms}. OverProt works with structures in
  \textbf{mmCIF format}. Structures deposited in the PDB {[}cite{]} are
  referenced by their PDB ID (e.g.~\texttt{1tqn}). OverProt follows the
  \emph{label*} numbering scheme when referencing chains and residues
  within a structure (i.e.~items \texttt{label\_asym\_id} and
  \texttt{label\_seq\_id} in the mmCIF file) -- this is sometimes
  different from the \emph{auth*} numbering scheme.
\item
  \textbf{Protein domain} -- a part of protein structure which may be a
  whole chain or a range (ranges) of residues in a chain. A domain is
  defined by the structure identifier, chain identifier, and one or more
  ranges of residues, e.g.~\texttt{1tqn,A,7:478} or
  \texttt{1n26,A,2:9,94:192}. Residue ranges include the start and end
  residue (e.g.~\texttt{5:8} means residues 5, 6, 7, 8).
\item
  \textbf{Protein family} -- a set of protein domains with reasonable
  structural similarity. The set can be provided by the user or it can
  be defined based on the CATH database {[}cite{]}, in which case the
  family (\emph{CATH superfamily}) is identified by its CATH ID
  (e.g.~\texttt{1.10.630.10}) and domains are identified by CATH domain
  ID (e.g.~\texttt{1tqnA00}).
\item
  \textbf{Secondary structure element (SSE)} -- a section of a protein
  chain with some secondary structure pattern. OverProt focuses on two
  types of SSEs -- \textbf{helices} (H) and \textbf{β-strands} (E). Each
  SSE within a protein structure can be identified by its chain
  identifier, start (index of its first residue), end (index of its last
  residue), and type. For comparing SSEs, it is convenient to simplify
  an SSE to a line segment (i.e.~3D coordinates of the start and end
  point).\\
  The term β-connectivity refers to the way in which the strands are
  connected: a \textbf{β-ladder} is a connection of two strands and can
  be either parallel or antiparallel; a \textbf{β-sheet} is a set of
  strands which are connected by β-ladders.\\
  This model is kept as simple as possible (different helix types
  (\(\alpha\), \(3_{10}\), \(\pi\)) are not distinguished; other SSE
  type (loops, turns) are not taken into account). Secondary structure
  assignment (i.e.~detection of SSEs) is performed by
  \textbf{SecStrAnnotator}, more details can be found in its original
  paper {[}cite{]}.
\item
  \textbf{Consensus SSE} -- a set of equivalent SSEs from different
  family members. The \textbf{occurrence} of a consensus SSE is the
  number of domains that contain it, divided by the total number of
  domains in the family. (e.g.~consensus helix X = \{helix H1 in
  domain1, helix H3 in domain2, helix H3 in domain3\}; domain4 contains
  no helix X; the occurrence is 3/4 = 0.75 or 75\%).
\item
  \textbf{Secondary structure consensus} -- a set of consensus SSEs
  together with their order and β-connectivity.
\end{itemize}

\hypertarget{methods-overprot-core}{%
\section{Methods -- OverProt Core}\label{methods-overprot-core}}

\textbf{OverProt Core} is an algorithm that constructs the secondary
structure consensus for a given protein family. The algorithm proceeds
in a number of steps, combined in \texttt{overprot.py}:

\hypertarget{preparation}{%
\subsection{Preparation}\label{preparation}}

\begin{itemize}
\item
  Download the list of domains for the family (if not already given by
  \texttt{-\/-domains}) from PDBe API
  (\texttt{https://www.ebi.ac.uk/pdbe/api/mappings/\{family\_id\}}).
\item
  Select sample: If \texttt{sample\_size} is different from
  \texttt{all}, select a random subset of the domain list.\\
  The family may contain multiple domains from the same PDB entry. If
  \texttt{{[}sample\_selection{]}unique\_pdb} is \texttt{True}, then
  these are treated as duplicates and only one of them is selected (the
  first in alphabetical order).
\item
  Download structures: The structures of listed domains are downloaded
  in mmCIF format, domains are cut out from the structures and saved in
  separate files. The sources of structures are given by
  \texttt{-\/-structure\_source} and
  \texttt{{[}download{]}structure\_sources}. Structures are also
  converted to PDB format for some later steps (MAPSCI). The download
  step is performed by \texttt{StructureCutter} written in C\#.
\end{itemize}

\hypertarget{structural-alignment}{%
\subsection{Structural alignment}\label{structural-alignment}}

Multiple structure alignment is performed in 2 steps:

\begin{itemize}
\item
  Program MAPSCI {[}cite Ilinkin et al., 2010{]} is used to calculate a
  consensus structure (-\textgreater{} \texttt{mapsci/consensus.cif}).
  For performance reasons, at most 100 domains are selected for this
  calculation (in a quasi-random way, i.e.~for the same family selects
  each time the same subset).\\
  To reduce indeterminism and ease later visualization, the consensus
  structure is centered to the origin (0, 0, 0), rotated so that its PCA
  components are aligned to the XYZ axes (``the structure is laid
  flat''), and flipped in a consistent way (roughly so that the chain
  goes left-top to right-bottom and its ends are more in front).\\
  In general, MAPSCI produces a reasonable consensus structure, but the
  alignment of the individual domains is often poor -- therefore the
  following re-alignment step is necessary.
\item
  In the realignment step, all domains are structurally aligned onto the
  consensus structure via \texttt{cealign} algorithm {[}cite Shindyalov
  and Bourne, 1998{]} provided in PyMOL module {[}cite{]}. In rare cases
  the cealign fail (when the domain is too short) -- in such cases a
  simple alignment algorithm \texttt{DumbAlign} is used instead
  (theretically inefficient and inaccurate, but sufficient for these
  very short domains).
\end{itemize}

\hypertarget{secondary-structure-assignment}{%
\subsection{Secondary structure
assignment}\label{secondary-structure-assignment}}

The SSEs in each domain are detected by \texttt{SecStrAnnotator} {[}cite
Midlik et al., 2019, 2021{]} (options \texttt{-\/-onlyssa}
\texttt{-\/-verbose} \texttt{-\/-batch}).

\hypertarget{guide-tree}{%
\subsection{Guide tree}\label{guide-tree}}

The domains are clustered by agglomerative clustering to produce a guide
tree. During this step, each domain is first converted into a
\textbf{weighted structure} (a sequence of C-alpha coordinates with
individual weight for each point). The two most similar weighted
structures are then merged into a new weighted structure, and this is
repeated until we end up with a single weighted structure, corresponding
to the tree root.

This agglomerative algorithm can be expressed by the following
pseudocode:

\begin{verbatim}
    Workset = { a weighted structure for each input domain }  
    while |Workset| > 1:  
        A, B = two nearest weighted structures in Workset
        C = merge(A, B)  
        Children[C] = {A, B}
        Workset = Workset - {A, B} ∪ {C}
\end{verbatim}

At the end, \texttt{Workset} will only contain one weighted structure,
which is the tree root. The topology of the tree will be defined by
\texttt{Children}.

Example: 5 structures were initially in \texttt{Workset}. B+C were
merged into BC, then D+E into DE, then A+BC into ABC, and finally ABC+DE
into ABCDE. The resulting tree is:

\begin{verbatim}
    A────────────┐
                 │
    B────┐       ├ABC──────┐
         ├BC─────┘         │
    C────┘                 ├ABCDE
                           │
    D────────┐             │
             ├DE───────────┘
    E────────┘
\end{verbatim}

Details of the algorithm are described in Appendix (distance function
\(D^{*}\), which determines the nearest weighted structures, operation
\texttt{merge}).

\hypertarget{merging}{%
\subsection{Merging}\label{merging}}

This is the core of the consensus generation algorithm. On the input, we
have a set of \emph{k} protein domains. Each domain is simplified to a
sequence of SSEs (defined by their type, line segment, etc.). The
required output is a clustering of all input SSEs. However, the
clustering must fulfil these constraints:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Each cluster can contain only elements of the same type (only helices
  or only strands).
\item
  A cluster must not contain more than one element from the same protein
  domain.
\item
  There must be a partial ordering of the clusters.
\end{enumerate}

The third constraint can be formalized:

\begin{itemize}
\tightlist
\item
  SSE \(x\) precedes SSE \(y\) (written \(x \rightarrow y\)) iff they
  are in the same protein domain and \(x\) appears before \(y\).
\item
  Cluster \(A\) precedes cluster \(B\) (\(A \rightarrow B\)) iff there
  exist SSEs \(x \in A, y \in B\) such that \(x \rightarrow y\).
\item
  There must be no sequence of clusters \(A, B...\), such that
  \(A \rightarrow B \rightarrow ... \rightarrow A\).
\end{itemize}

TODO figures (from presentations/posters) - GM pr16

In the merging step, the guide tree leaves are first populated with SSEs
of their respective domains. In each internal node, the SSEs from the
two children nodes are matched together and merged. The root then
contains the consensus SSEs of the whole family. The matching and
merging of two SSE graphs is in principle similar to matching (aligning)
and merging of two weighted structures. The best matching is also found
by dynamic programming. However, it is more complicated here, because 1)
clusters of different type cannot be matched (this can cause the
branching in the resulting graph), and 2) the dynamic programming
algorithm is not as straightforward for matching DAGs as it is for
matching sequences.

TODO say what is implemented in which function

\hypertarget{annotation}{%
\subsection{Annotation}\label{annotation}}

This is an optional step, in which the generated SSE consensus is used
as annotation template for SecStrAnnotator and all family members are
annotated. Before the annotation, the SSEs with low occurrence
(\textless{} 5\%) are removed, which dramatically reduces the running
time of SecStrAnnotator. Options for SecStrAnnotator:
\texttt{-\/-ssa\ file} ~\texttt{-\/-align\ none}
~\texttt{-\/-metrictype\ 3} ~\texttt{-\/-fallback\ 30}
~\texttt{-\/-unannotated}.

\hypertarget{visualization}{%
\subsection{Visualization}\label{visualization}}

The generated SSE consensus is visualized by several SVG diagrams with
different settings and \texttt{diagram.json} file is produced, which
will be used for interactive visualization be OverProt Viewer. A PyMOL
session (.pse) is created, with the structure consensus from MAPSCI
(shown as ribbon) and the SSE consensus (shown as flat-end cylinders
(helices) and round-end cylinders (strands)). A PNG image is also
rendered from the session. A session with all domains and their SSEs is
generated if \texttt{{[}visualization{]}create\_multi\_session} is
\texttt{True} (very slow, not recommended for larger families).

\hypertarget{multi-family-mode}{%
\subsection{Multi-family mode}\label{multi-family-mode}}

Multiple families can be processed in parallel using
\texttt{overprot\_multifamily.py}.

\hypertarget{interactive-visualization-by-overprot-viewer}{%
\section{Interactive visualization by OverProt
Viewer}\label{interactive-visualization-by-overprot-viewer}}

\textbf{OverProt Viewer} is a web component for interactive
visualization of the SSE consensus. Its input is the preprocessed
\texttt{diagram.json} file. It is implemented in TypeScript with D3.js.

\hypertarget{data-computation-for-overprot-server}{%
\section{Data computation for OverProt
Server}\label{data-computation-for-overprot-server}}

\textbf{OverProt Server} serves precomputed SSE consensuses (database)
and runs the OverProt Core algorithm for user-defined sets of domains
(jobs).

The database is constructed in this way:

\begin{itemize}
\tightlist
\item
  Retrieve the current list of families from CATH
  (ftp://orengoftp.biochem.ucl.ac.uk\allowbreak/cath/releases/latest-release/cath-classification-data/cath-domain-list.txt).
  (This file also contains the domain definitions, but they are in an
  incompatible numbering scheme (\emph{auth*}), therefore they are not
  used). This is currently 6631 families, out of which 64 are empty
  families (December 2021).
\item
  Retrieve the domain lists for each family, including chains and
  residue ranges, from PDBe API
  (https://www.ebi.ac.uk/pdbe/api/mappings/\{family\_id\}). This is
  currently over 470k domains (December 2021).
\item
  Remove duplicates (i.e.~multiple domains from the same PDB entry).
  This is currently over 200k domains (December 2021).
\item
  Apply the OverProt Core algorithm to each family.
\end{itemize}

The whole process is realized by:

\begin{Shaded}
\begin{Highlighting}[]
    \ExtensionTok{python}\NormalTok{  overprot_multifamily.py  --download_family_list_by_size \textbackslash{}}
\NormalTok{    --config working_scripts/overprot-config-overprotserverdb.ini \textbackslash{}}
\NormalTok{    --collect  xxx  all  }\VariableTok{$UPDATE_DIRECTORY}  
\end{Highlighting}
\end{Shaded}

The database is updated weekly.

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

\hypertarget{computation-of-distance-function-for-two-weighted-structures}{%
\subsection{Computation of distance function for two weighted
structures}\label{computation-of-distance-function-for-two-weighted-structures}}

A weighted structure \(A\) is a tuple
\((n^A, \mathbf{R}^A, \mathbf{W}^A, k^A)\) where \(n^A\) is the length
of the weighted structure (number of points), \(\mathbf{R}^A\) is the
matrix of their coordinates (\(n^A \times 3\)), \(\mathbf{W}^A\) is the
vector of their relative weights \(\in (0,1]\), and \(k^A\) is the
absolute weight of \(A\) (scalar). Example of a weighted structure:

\[n^A = 4 \quad
\mathbf{R}^A = \begin{bmatrix}-1.1&-2.9&0.1&0.4\\0.0&1.1&0.9&-2.7\\5.2&2.1&0.0&0.8\end{bmatrix} \quad
\mathbf{W}^A = \begin{bmatrix}1&0.5&0.8&1\end{bmatrix} \quad
k^A = 10\]

\(\mathbf{r}^A_i\) and \(w^A_i\) will refer to \(i\)-th column of
\(\mathbf{R}^A\) and \(\mathbf{W}^A\).

TODO nice format (matrices and vectors in bold)

A protein domain can be converted into a weighted structure as follows:
\(n\) is the number of residues, \(r^A_i\) are the coordinates of the
C-alpha atom of \(i\)-th residue, \(w^A_i\) is 1, and \(k^A\) is 1.

Distance function \(d\) is defined for two weighted points:

\[d((r^A_i, w^A_i), (r^B_j, w^B_j)) = \left(1 - e^{-\lVert r^A_i-r^B_j \rVert / R_0}\right) \cdot \min\{ w^A_i, w^B_j \} + \frac{1}{2} \lvert w^A_i-w^B_j \rvert\]

In case that one of the weighted points is undefined (\(\bot\)), \(d\)
is still defined:

\[d((r^A_i, w^A_i), \bot) = \frac{1}{2} w^A_i \qquad d(\bot, (r^B_j, w^B_j)) = \frac{1}{2} w^B_j\]

(\(d\) is not the Euclidean distance of the two points.
\(d \in [0,1)\).)

An alignment of two weighted structures \(A\), \(B\) is a sequence of
pairs \([(p_1, q_1), (p_2, q_2), ..., (p_n, q_n)]\) (for better reading
written as a matrix
\(\begin{bmatrix}p_1&p_2&...&p_n\\q_1&q_2&...&q_n\end{bmatrix}\)), where
\(p_i\) and \(q_i\) are indices of the points of \(A\) and \(B\).
Indices must be increasing and must include each index exactly once for
both \(A\) and \(B\). Value \(\bot\) means that a particular point was
not matched. Example of a valid alignment for \(n^A = 4, n^B = 5\):

\[\begin{bmatrix}1&2&3&4&\bot&\bot\\1&\bot&2&3&4&5\end{bmatrix}\]

\(D(A, B, M) = \sum\limits_{(p, q) \in M}{d((r^A_{p}, w^A_{p}), (r^B_{q}, w^B_{q}))}\)

The best alignment \(M^*(A, B)\) is the alignment \(M\) which minimizes
\(D(A, B, M)\).

The distance of \(A\) and \(B\) is \(D^*(A, B) = D(A, B, M^*(A, B))\).

The best alignment is found by dynamic programming (this is basically
the Needleman-Wunsch algorithm {[}cite{]}) -- for this, the distance
function \(d\) is converted into a score function \(s\):

\[s((r^A_i, w^A_i), (r^B_j, w^B_j)) = \frac{1}{2} w^A_i + \frac{1}{2} w^B_j - d((r^A_i, w^A_i), (r^B_j, w^B_j))\]

\[s((r^A_i, w^A_i), \bot) = 0 \qquad s(\bot, (r^B_j, w^B_j)) = 0\]

\[S(A, B, M) = \sum\limits_{(p, q) \in M}{s((r^A_{p}, w^A_{p}), (r^B_{q}, w^B_{q}))}\]

\[S(A, B, M) = \sum\limits_{i=1}^{n^A}{w^A_i} + \sum\limits_{j=1}^{n^B}{w^B_j} - D(A, B, M)\]

From this equation it can be seen, that maximizing \(S\) by dynamic
programming also minimizes \(D\).

A useful property of the distance function \(D^*\) is that it is a
metric (i.e.~\(D^*(A,A) = 0\), \(D^*(A,B) = D^*(B,A)\), and
\(D^*(A,B) + D^*(B,C) \geq D^*(A,C)\) for any weighted structures
\(A, B, C\)).

Having two structures \(A, B\) and their best alignment
\(M^*(A, B) = [(p_1, q_1), ..., (p_n, q_n)]\), we can define operation
\(merge\) as follows:

\[merge(A, B) = C = (n^C, R^C, W^C, k^C)\]

\[n^C = n\]

\[r^C_i = \frac {r^A_{p_i} w^A_{p_i} k^A + r^B_{q_i} w^B_{q_i} k^B} {w^A_{p_i} k^A + w^B_{q_i} k^B}\]

\[w^C_i = w^A_{p_i} k^A + w^B_{q_i} k^B\]

\[k^C = k^A + k^B\]

(If \(p_i=\bot\), the values can be calculated by setting
\(w^A_{p_i} = 0\), thus simplifying to
\(r^C_i = r^B_{q_i}, w^C_i = w^B_{q_i}\). Similarly for \(q_i=\bot\).)

Remark: When finding the two nearest items in the work set, it is not
necessary to calculate the distance \(D^*\) for every pair of items --
there are specialized data structures that can significantly decrease
the number of distance calculations. We use a non-standard structure
NN-tree (nearest neighbour tree). (Standard structures like VP-tree,
GH-tree, GNAT, M-tree etc. (TODO name, what I tried) either miss some of
the necessary operations (insert, delete) or perform worse than NN-tree
for this particular application.) In some larger protein families this
can reduce the number of distance computations to roughly 20\%.

\hypertarget{references}{%
\section{References}\label{references}}

\href{http://example.com/}{Exemplini F \emph{et al.} (2019) Examples in
scientific literature. J Exemp Biol, 42, 1-100.}

\end{document}
